---
title: "Learn R through Swirl"
author: "Rohit Padebettu"
date: "5/26/2017"
output:
  pdf_document: default
  html_document: default
---
## Introduction
This tutorial is a **necessary first step** for students who have had no past experience in using R. It is recommended that students complete the first course *"R Programming"*, which is a collection of 15 sub-modules before working on the first case study. This is a collection of interactive courses for use with the [swirl R package](http://swirlstats.com). You'll find instructions for installing courses below on this page.

## Pre-requisites
Pre-requisites for this tutorial are to have access to a working version of RStudio installed on your machine. Please follow and complete the RStudio installation tutorial before getting started on this.

## Install and run a course automatically from swirl

**This is the preferred method of installing courses.** It automates the process by allowing you to do everything right from the R console.

1) Make sure you have a recent version version of swirl:

```{r, eval=F}
install.packages("swirl")
```

2) Enter the following from the R console, **substituting the name of the course** that you wish to install:

```{r, eval=F}
library(swirl)
install_course("Course Name Here")
swirl()
```

For example, `install_course("R Programming")` will install the R Programming course. **Please note that course names are case sensitive!**

If that doesn't work for you...

## Install and run a course manually

If the automatic course installation method outlined above does not work for you, then there's a simple alternative.

1) Click [**here**](https://github.com/swirldev/swirl_courses/archive/master.zip) to download the file `swirl_courses-master.zip`.

2) Enter the following from the R console, **substituting the correct file path** to your downloaded file and the **name of your desired course**:

```
library(swirl)
install_course_zip("path/to/file/here/swirl_courses-master.zip", multi=TRUE, 
                   which_course="Course Name Here")
swirl()
```

For example, if you download the zip file to `~/Downloads/swirl_courses-master.zip`, then the following command will install the R Programming course.

```
install_course_zip("~/Downloads/swirl_courses-master.zip", multi=TRUE, which_course="R Programming")
```

**Please note that course names are case sensitive!**

Although we recommend you install one course at a time, if you omit the `which_course` argument, then all available courses from this repository will be installed:

```
install_course_zip("~/Downloads/swirl_courses-master.zip", multi=TRUE)
```

## R Programming
The R Programming course is a collection of 15 sub-modules to help familiarize you with some basic operations done in R. the modules are as listed below

 1. Basic Building Blocks
 2. Workspace and Files
 3. Sequences of Numbers
 4. Vectors             
 5. Missing Values 
 6. Subsetting Vectors
 7. Matrices and Data Frames
 8. Logic            
 9. Functions
 10. lapply and sapply 
 11. vapply and tapply
 12. Looking at Data       
 13. Simulation
 14. Dates and Times
 15. Base Graphics


## Uninstall a course

If you'd like to remove a course at any time, you can use `uninstall_course("Course Name Here")`.
